Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-07-07 paper/arm.tex
Minor spelling and editorial changes

  Execution $\aEG$ is
  \emph{$({\rco}, {\rrfx}, {\rcb})$-valid} under \emph{External Consistency} (\EC{}) if
  \begin{enumerate}[resume,label=(\textsc{a}\arabic*),ref=\textsc{a}\arabic*]
  \item[\eqref{arm-co}] and \eqref{arm-rf}, as for \EGC,
  \item \label{arm-cb}
    ${\rcb}\supseteq\PBR{{\rco}\cup{\rlob}}$ is a linear order %, capturing \emph{completes before}, %$\rcb : % \aEvs\times\aEvs$
    such that if $\bEv\xrfx\aEv$ then either
    \begin{enumerate}%[leftmargin=0pt]
    \item \label{arm-rfe}
      % $\bEv\xcb\aEv$ and $\cEv\rblocks\aEv$ then either $\cEv\xcb\bEv$ or $\aEv\xcb\cEv$, or 
      $\bEv\xcb\aEv$ and if $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\xcb\bEv$ or $\aEv\xcb\cEv$, or
      % $\bEv\xcb\aEv$ and $(\not\exists\cEv)$ such that $\bEv\xcb\cEv\xcb\aEv$ and $\cEv\rblocks\aEv$, or 
    \item \label{arm-rfi}
      $\bEv\xcbinv\aEv$ and $\bEv\xpoloc\aEv$ and $(\not\exists\cEv)$ $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv\xpoloc\cEv\xpoloc\aEv$.
      % $\bEv\xpoloc\aEv$ and $(\not\exists\cEv)$ such that $\bEv\xpoloc\cEv\xpoloc\aEv$ and $\cEv\rblocks\aEv$.
      % \item if $\bEv\xco\aEv$ then $\bEv\xgcb\aEv$,
      % \item if $\bEv\xrfx\aEv$ then $\bEv\xgcb\aEv$,
      % \item if $\bEv\xrfx\aEv$ and $\cEv\rblocks\aEv$ then either $\cEv\xgcb\bEv$ or $\aEv\xgcb\cEv$,
      % \item if $\bEv\xlob\aEv$ then either $\bEv\xgcb\aEv$ or $(\exists\cEv)$
      %   $\cEv\xrfx\bEv$ and $\cEv\xpoloc\bEv$ but not $\cEv\xlob\aEv$.
    \end{enumerate}
  \end{enumerate}
\end{definition}
\citet{armed} show that \EGC{} and \EC{} are both equivalent to the standard
definition of Arm8.  They explain \EGC{} and \EC{} using the following
example, which is allowed by Arm8.\footnote{We have changed an address
  dependency in the first thread to a data dependency.}
\begin{gather*}
  \PW{x}{1}\SEMI 
  \PR{x}{r}\SEMI
  \PW{y}{r} \PAR
  \PR[\mACQ]{y}{1}\SEMI
  \PR{x}{s}
  \\
  %\tag{\cmark\armeight}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \rfx{a}{b}
      \lob{b}{c}
      \rfx{c}{d}
      \lob{d}{e}
      \fr[out=-165,in=-15]{e}[above,pos=.375]{a}
    \end{tikzinline}}
\end{gather*}
\EGC{} drops $\rlob$-order in the first thread using \ref{arm-gcb-lob}, since
$\DWP{x}{1}$ is not $\rlob$-ordered before $\DWP{y}{1}$.
\begin{gather*}
  \tag{$\rgcb$}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \gcbz{a}{b}
      \gcbz{c}{d}
      \gcbz{d}{e}
      \gcbz[out=-165,in=-15]{e}{a}
      % \rfx{a}{b}
      % %\lob{b}{c}
      % \rfx{c}{d}
      % \lob{d}{e}
      % \co[out=-165,in=-15]{e}[above right]{a}
    \end{tikzinline}}
\end{gather*}
\EC{} drops $\rrfx$-order in the first thread using \ref{arm-rfi}.
\begin{gather*}
  \tag{$\rcb$}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \cbz{b}{c}
      \cbz{c}{d}
      \cbz{d}{e}
      \cbz[out=-165,in=-15]{e}{a}
      % %\rfx{a}{b}
      % \lob{b}{c}
      % \rfx{c}{d}
      % \lob{d}{e}
      % \co[out=-165,in=-15]{e}[above right]{a}
    \end{tikzinline}}
\end{gather*}

>>>>

  Execution $\aEG$ is
  \emph{$({\rco}, {\rrfx}, {\rcb})$-valid} under \emph{External Consistency} (\EC{}) if
  \begin{enumerate}[resume,label=(\textsc{a}\arabic*),ref=\textsc{a}\arabic*]
  \item[\eqref{arm-co}] and \eqref{arm-rf}, as for \EGC,
  \item \label{arm-cb}
    ${\rcb}\supseteq\PBR{{\rco}\cup{\rlob}}$ is a linear order %, capturing \emph{completes before}, %$\rcb : % \aEvs\times\aEvs$
    such that if $\bEv\xrfx\aEv$ then either
    \begin{enumerate}%[leftmargin=0pt]
    \item \label{arm-rfe}
      % $\bEv\xcb\aEv$ and $\cEv\rblocks\aEv$ then either $\cEv\xcb\bEv$ or $\aEv\xcb\cEv$, or 
      $\bEv\xcb\aEv$ and if $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\xcb\bEv$ or $\aEv\xcb\cEv$, or
      % $\bEv\xcb\aEv$ and $(\not\exists\cEv)$ such that $\bEv\xcb\cEv\xcb\aEv$ and $\cEv\rblocks\aEv$, or 
    \item \label{arm-rfi}
      $\bEv\xcbinv\aEv$ and $\bEv\xpoloc\aEv$ and $(\not\exists\cEv)$ $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv\xpoloc\cEv\xpoloc\aEv$.
      % $\bEv\xpoloc\aEv$ and $(\not\exists\cEv)$ such that $\bEv\xpoloc\cEv\xpoloc\aEv$ and $\cEv\rblocks\aEv$.
      % \item if $\bEv\xco\aEv$ then $\bEv\xgcb\aEv$,
      % \item if $\bEv\xrfx\aEv$ then $\bEv\xgcb\aEv$,
      % \item if $\bEv\xrfx\aEv$ and $\cEv\rblocks\aEv$ then either $\cEv\xgcb\bEv$ or $\aEv\xgcb\cEv$,
      % \item if $\bEv\xlob\aEv$ then either $\bEv\xgcb\aEv$ or $(\exists\cEv)$
      %   $\cEv\xrfx\bEv$ and $\cEv\xpoloc\bEv$ but not $\cEv\xlob\aEv$.
    \end{enumerate}
  \end{enumerate}
\end{definition}
\citet{armed} show that \EGC{} and \EC{} are both equivalent to the standard
definition of \armeight.  They explain \EGC{} and \EC{} using the following
example, which is allowed by \armeight.\footnote{We have changed an address
  dependency in the first thread to a data dependency.}
\begin{gather*}
  \PW{x}{1}\SEMI 
  \PR{x}{r}\SEMI
  \PW{y}{r} \PAR
  \PR[\mACQ]{y}{1}\SEMI
  \PR{x}{s}
  \\
  %\tag{\cmark\armeight}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \rfx{a}{b}
      \lob{b}{c}
      \rfx{c}{d}
      \lob{d}{e}
      \fr[out=-165,in=-15]{e}[above,pos=.375]{a}
    \end{tikzinline}}
\end{gather*}
\EGC{} drops $\rlob$-order in the first thread using \ref{arm-gcb-lob}, since
$\DWP{x}{1}$ is not $\rlob$-ordered before $\DWP{y}{1}$.
\begin{gather*}
  \tag{$\rgcb$}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \gcbz{a}{b}
      \gcbz{c}{d}
      \gcbz{d}{e}
      \gcbz[out=-165,in=-15]{e}{a}
      % \rfx{a}{b}
      % %\lob{b}{c}
      % \rfx{c}{d}
      % \lob{d}{e}
      % \co[out=-165,in=-15]{e}[above right]{a}
    \end{tikzinline}}
\end{gather*}
\EC{} drops $\rrfx$-order in the first thread using \ref{arm-rfi}.
\begin{gather*}
  \tag{$\rcb$}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \cbz{b}{c}
      \cbz{c}{d}
      \cbz{d}{e}
      \cbz[out=-165,in=-15]{e}{a}
      % %\rfx{a}{b}
      % \lob{b}{c}
      % \rfx{c}{d}
      % \lob{d}{e}
      % \co[out=-165,in=-15]{e}[above right]{a}
    \end{tikzinline}}
\end{gather*}

### 2021-07-07 paper/intro.tex
Minor spelling and editorial changes

Recently, two denotational models have been proposed that compute sequential
dependencies semantically. \citet{DBLP:conf/esop/PaviottiCPWOB20} defined
Modular Relaxed Dependencies (\MRD), which use event structures to calculate
dependencies for \cXI.  \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined
Pomsets with Preconditions (\PwP), which use preconditions and logic to
calculate dependencies for a Java-like language on \mca{} hardware, such as
\armeight{} \cite{DBLP:journals/pacmpl/PulteFDFSS18}.  However, neither paper
treated sequential composition as a first-class citizen.  \MRD{} encoded
sequential composition using continuation-passing, and \PwP{} used prefixing,
adding one event at a time on the left.  In both cases, the approaches
require perfect knowledge of the future.

@@

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order of reads and writes (black arrows),
where each read has a matching \emph{fulfilling} write (green arrows),
for example:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}\SEMI
    \PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \po{wx0}{wx1}
      \po{wx1}{wy2}
      \po{wy2}{ry2}
      \po{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}

>>>>

Recently, two denotational models have been proposed that compute sequential
dependencies semantically. \citet{DBLP:conf/esop/PaviottiCPWOB20} defined
Modular Relaxed Dependencies (\MRD), which use event structures to calculate
dependencies for \cXI.  \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined
Pomsets with Preconditions (\PwP), which use preconditions and logic to
calculate dependencies for a Java-like language on multicopy-atomic (\mca{}) hardware, such as
\armeight{} \cite{DBLP:journals/pacmpl/PulteFDFSS18}.  However, neither paper
treated sequential composition as a first-class citizen.  \MRD{} encoded
sequential composition using continuation-passing, and \PwP{} used prefixing,
adding one event at a time on the left.  In both cases, the approaches
require perfect knowledge of the future.

@@

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order of reads and writes (pink arrows \tikz[baseline]{\draw [po] (0,0.1) -- (0.5,0.1);}),
where each read has a matching \emph{fulfilling} write (green arrows \tikz[baseline]{\draw [rf] (0,0.1) -- (0.5,0.1);}),
for example:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}\SEMI
    \PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \po{wx0}{wx1}
      \po{wx1}{wy2}
      \po{wy2}{ry2}
      \po{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}
